Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-21 paper.tex
progress

\noindent Several related topics are not handled in this paper.

\begin{itemize}

\item
Besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. However, we
do keep canonicity in mind while specifying our systems, and aim to avoid
pitfalls which would cause canonicity to fail.
\item
We only focus on using universes as sized-based classifiers for
types. Universe-like stratification is also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}, but we do not attempt to cover such use
cases.
\item
We only cover predicative hierarchies.
\item
We do not discuss ergonomics and implementation of elaboration. Our goal is to
develop a system which is as flexible as possible, even though that may not be
the best suited for inference and elaboration. For example, first-class levels
have undecidable ordering in general.
\end{itemize}

@@

\begin{mydefinition} A model of TTDL consists of the following.
\begin{enumerate}
\item
  A base category ($\Con$, $\Sub$) with terminal object $\emptycon$.
\item
  A ``dependent'' level structure on the base category:
  \begin{alignat*}{3}
    &\Lvl                  &&: \Con \to \Seti\\
    &\blank\!<\!\blank     &&: \{\Gamma : \Con\} \to \Lvl\,\Gamma \to \Lvl\,\Gamma \to \Seti\\
    &<\!\msf{prop}         &&: (p\,q : i < j) \to p = q\\
    &\blank\!\circ\!\blank &&: j < k \to i < j \to i < k
  \end{alignat*}
  Additionally, $\Lvl$ and $\blank\!<\!\blank$ are natural in the base category,
  so they support substitution operations. \emph{Remark:} at this point, we do not require well-foundedness
  for $\blank\!<\!\blank$, as it has no bearing on basic lifting and universe rules, and
  well-foundedness will be usually internally provable when we add level reflection rules.
\item
  A ``bootstrapping'' assumption on levels. This can be any collection of levels
  and morphisms. We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist
  together with $l_{01} : l_0 < l_1$. We need to assume at least some levels,
  because otherwise the TTDL syntax is trivial; it is not even possible to
  write down a non-empty context or a closed type former. $l_0 < l_1$ allows large
  eliminations on type formers so it provides a fair amount of expressive power.
\item
  A family structure:
  \begin{alignat*}{3}
    &\Ty &&: (\Gamma : \Con) \to \Lvl\,\Gamma \to \Seti\\
    &\Tm &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Seti\\
    &\blank\ext\blank &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Con
  \end{alignat*}
  We have type and term substitution, which depends on level substitution. For instance, we have
  \[
    \blank[\blank] : \Ty\,\Delta\,i \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Ty\,\Gamma\,(i[\sigma])
  \]
  We also have a comprehension isomorphism $\Sub\,\Gamma\,(\Delta \ext A) \simeq
  (\sigma : \Sub\,\Gamma\,\Delta)\times \Tm\,\Gamma\,(A[\sigma])$, which is
  natural in $\Gamma$.
\item A lifting structure with
  \begin{alignat*}{3}
    &\Lift &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\} \to i < j \to \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j\\
    &\up   &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\}(p : i < j) \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,(\Lift\,p\,A)
  \end{alignat*}
  Such that
  \vspace{1em}
  \begin{enumerate}
    \item $\Lift$ preserves all basic type formers and has functorial action on $p \circ q$.
    \item $\up$ has an inverse $\down$, preserves all basic term formers and has functorial action on $p \circ q$.
    \item $(\Gamma \ext A) = (\Gamma \ext \Lift\,p\,A)$, and $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$ and $\up\,t = t$.
    %% \item $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$.
    %% \item $\up\,t = t$.
  \end{enumerate}
  \vspace{1em}
  Above we mention basic type formers, although we haven't yet specified
  those. The way this should be understood, is that any basic type former introduced
  from now on should come equipped with preservation equations for lifting. This is
  similar to how we mandate that any introduced type former must be natural with respect
  to substitution.
\item A universe structure
  \begin{alignat*}{6}
    &\U  &&: \{\Gamma : \Con\}(i\,j : \Lvl\,\Gamma) \to i < j \to \Ty\,\Gamma\,j\hspace{2em}
    &\El &&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
  \end{alignat*}
  Such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
  inverse $\Code$, $\Tm\,\Gamma\,(\U\,i\,j\,p) = \Ty\,\Gamma\,i$ and $\El\,t =
  t$.
\item Basic type formers.
\end{enumerate}
\end{mydefinition}

@@

\section{Future Work and Conclusion} \label{sec:conclusion}

>>>>

@@

\begin{mydefinition} A model of TTDL consists of the following.
\begin{enumerate}
\item
  A base category ($\Con$, $\Sub$) with terminal object $\emptycon$.
\item
  A ``dependent'' level structure on the base category:
  \begin{alignat*}{3}
    &\Lvl                  &&: \Con \to \Seti\\
    &\blank\!<\!\blank     &&: \{\Gamma : \Con\} \to \Lvl\,\Gamma \to \Lvl\,\Gamma \to \Seti\\
    &<\!\msf{prop}         &&: (p\,q : i < j) \to p = q\\
    &\blank\!\circ\!\blank &&: j < k \to i < j \to i < k
  \end{alignat*}
  Additionally, $\Lvl$ and $\blank\!<\!\blank$ are natural in the base category,
  so they support substitution operations. \emph{Remark:} at this point, we do not require well-foundedness
  for $\blank\!<\!\blank$, as it has no bearing on basic lifting and universe rules, and
  well-foundedness will be usually internally provable when we add level reflection rules.
\item
  A ``bootstrapping'' assumption on levels. This can be any collection of levels
  and morphisms. We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist
  together with $l_{01} : l_0 < l_1$. We need to assume at least some levels,
  because otherwise the TTDL syntax is trivial; it is not even possible to
  write down a non-empty context or a closed type former. $l_0 < l_1$ allows large
  eliminations on type formers so it provides a fair amount of expressive power.
\item
  A family structure:
  \begin{alignat*}{3}
    &\Ty &&: (\Gamma : \Con) \to \Lvl\,\Gamma \to \Seti\\
    &\Tm &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Seti\\
    &\blank\ext\blank &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Con
  \end{alignat*}
  We have type and term substitution, which depends on level substitution. For instance, we have
  \[
    \blank[\blank] : \Ty\,\Delta\,i \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Ty\,\Gamma\,(i[\sigma])
  \]
  We also have a comprehension isomorphism $\Sub\,\Gamma\,(\Delta \ext A) \simeq
  (\sigma : \Sub\,\Gamma\,\Delta)\times \Tm\,\Gamma\,(A[\sigma])$, which is
  natural in $\Gamma$.
\item A lifting structure with
  %% \vspace{-0.5em}
  \begin{alignat*}{3}
    &\Lift &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\} \to i < j \to \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j\\
    &\up   &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\}(p : i < j) \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,(\Lift\,p\,A)
  \end{alignat*}
  Such that
  \begin{enumerate}
    \item $\Lift$ preserves all basic type formers and has functorial action on $p \circ q$.
    \item $\up$ has an inverse $\down$, preserves all basic term formers and has functorial action on $p \circ q$.
    \item $(\Gamma \ext A) = (\Gamma \ext \Lift\,p\,A)$, and $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$ and $\up\,t = t$.
    %% \item $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$.
    %% \item $\up\,t = t$.
  \end{enumerate}

  Above we mention basic type formers, although we haven't yet specified
  those. The way this should be understood, is that any basic type former introduced
  from now on should come equipped with preservation equations for lifting. This is
  similar to how we mandate that any introduced type former must be natural with respect
  to substitution.
\item A universe structure
  %% \vspace{-0.5em}
  \begin{alignat*}{6}
    &\U  &&: \{\Gamma : \Con\}(i\,j : \Lvl\,\Gamma) \to i < j \to \Ty\,\Gamma\,j\hspace{2em}
    &\El &&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
  \end{alignat*}
  Such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
  inverse $\Code$, $\Tm\,\Gamma\,(\U\,i\,j\,p) = \Ty\,\Gamma\,i$ and $\El\,t =
  t$.
\item Basic type formers.
\end{enumerate}
\end{mydefinition}

@@

\section{Conclusion and Future Work} \label{sec:conclusion}

In the current work, we developed a framework for modeling a variety of universe
features in type theories. At this point, we may ask the question: if IR is
sufficient to model every feature, why not simply add IR support in a practical
implementation, and drop the menagerie of universe features?

The answer is that IR provides is a \emph{deep embedding} of universe features,
which is usually less convenient to use as \emph{native} features. For example,
both Coq and Agda have powerful automatic solving for filling out implicit
universe levels. We also do not have to invoke $\El$ or the $\ult$ computation
rule explicitly, and in Coq we can use implicit syntax for subtyping instead of
explicit coercions.

This trade-off between convenience and formal minimalism is similar to the
situation with inductive types. Formally, W-types and identity types are easier
to handle than general inductive families, but the latter are far more
convenient to actually use. Ideally, we would like to justify complicated
convenience features by reduction to minimal features. With the current paper,
we hope to have made progress in this manner.

\subsection{Future Work}

Several related are not discussed in this paper and could be subject to future
work.

First, besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. We did keep
canonicity in mind when specifying the systems in this paper. Hopefully the
usual proof method of gluing (in other words, proof-relevant logical predicates)
\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
this paper. It seems that the main task in doing so is to define proof-relevant
logical predicates over $\uir$.

Second, we only focus on using universes as sized-based classifiers for
types. Stratification features are also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}. It would be interesting to port
universe features in this paper to two-level type theory, as they would
hopefully model a form of stage polymorphism in multi-stage compilation.

Third, we do not discuss implementation strategies and ergonomics of universe
features. Which universe hierarchies support good proof automation? What kind of
impact do first-class levels have on elaboration algorithms? Hopefully the
current work can aid answering these questions, by at least giving a way to quickly
check if some features are logically consistent.

Lastly, we do not handle impredicative universes. The main reason for this is
that we do not know the consistency of having IR and impredicative function
space together in the same universe, and modeling impredicativity seems to
require this assumption in the metatheory. This could be investigated as well
in future research.
